---
title: "Getting started with synthsdtm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with synthsdtm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # The pipeline writes to disk and resolves Controlled Terminology from the
  # bundled snapshot. We show the calls and their expected (deterministic)
  # output rather than running them while the vignette builds.
  eval = FALSE
)
```

`synthsdtm` generates synthetic, CDISC SDTMIG 3.4-shaped clinical-trial datasets
from a single machine-readable study configuration. The guiding idea is that the
**study is data, not code**: cohorts, design, the visit grid, test panels, and
rate knobs all live in a `study_config.json`. One config-driven engine turns that
config into SDTM datasets, so adapting to a new protocol means writing a new
config — never editing R.

Generation is **deterministic and seeded**: the same config, spec, CT, subject
count, and seed always produce identical output. Shape-domain row counts (DM, IE,
EX, DS, and the findings domains) are a closed-form function of the structural
inputs and do not depend on the seed; only value cells vary, and they stay within
documented bounds.

```{r install}
# from the package directory
devtools::install(".")        # or devtools::load_all(".") during development
library(synthsdtm)
```

`jsonlite` is required; `haven` is needed only for XPT export.

## The pipeline at a glance

```
1. PLAN       study_config.json  +  build_spec()  → spec + pinned CT
2. GENERATE   simulate_sdtm(config, spec, ct, n, seed)  → SDTM data.frames
3. CHECK      check_sdtm()  → T1 Shape · T2 Conformance · T3 Cross-domain  (gate)
              export_xpt() + digest_core()  → XPT v5 + CDISC CORE          (optional)
```

The package bundles a reference config for **NCT04556760** that we use throughout.

```{r config}
cfg <- system.file("configs", "NCT04556760.json", package = "synthsdtm")
```

## 1. Resolve the spec and Controlled Terminology

`build_spec()` reads the config, selects the SDTMIG 3.4 variable spec for the
in-scope domains, and pins Controlled Terminology from the bundled offline
snapshot (no API key, fully reproducible). Passing `out_dir` also writes
`sdtm_spec.json`, `ct_cache.json`, and `coverage.json`.

```{r build-spec}
built <- build_spec(cfg, out_dir = "runs/ref/config")
names(built)
#> [1] "spec"      "ct_cache"  "coverage"
```

If the config names a domain with no SDTMIG template, `build_spec()` fails fast
and points you at `supported_domains()` (and the Findings-class shortcut) rather
than silently dropping it.

## 2. Generate the SDTM datasets

`simulate_sdtm()` returns an `sdtm_sim` object — one `data.frame` per domain,
plus lineage and summary metadata. `write_sdtm()` emits one `<DOMAIN>.csv` per
domain along with the CORE metadata sidecars (`_datasets.csv`, `_variables.csv`),
`lineage.json`, and `datasets_summary.json`.

```{r simulate}
res <- simulate_sdtm(cfg, built$spec, built$ct_cache,
                     n_subjects = 40, seed = 1234)
res$summary
#> $DM 40   $IE 2   $VS 720   $EG 640   $LB 2080   $EX 80   $DS 40   ...

head(res$domains$DM[, c("USUBJID", "ARMCD", "ARM", "AGE", "SEX")])

write_sdtm(res, "runs/ref/sdtm")
```

`n_subjects` overrides the cohort sizes in the config (the cohort split is scaled
to the requested total); omit it to use the config's own `Σn`. The seed defaults
to `1234`.

These row counts are exact and seed-independent for this config:

| DM | IE | VS  | EG  | LB   | EX | DS |
|----|----|-----|-----|------|----|----|
| 40 | 2  | 720 | 640 | 2080 | 80 | 40 |

The seed-dependent domains (MH, CM, AE) land within documented bounds.

## 3. Check the output

`check_sdtm()` recomputes the expected shape from the config and asserts on the
data written to `run_dir`, across three tiers:

- **T1 Shape** — exact row counts per domain, recomputed from the structural inputs.
- **T2 Conformance** — variable presence/order, types, and Controlled Terminology.
- **T3 Cross-domain** — referential and temporal consistency between domains.

```{r check}
rep <- check_sdtm(cfg, built$spec, built$ct_cache,
                  run_dir = "runs/ref/sdtm", n_subjects = 40, seed = 1234)
#> Accuracy checks: 126/126 passed
#>   T1: .../... pass
#>   T2: .../... pass
#>   T3: .../... pass
#> OVERALL: ALL PASS

stopifnot(rep$allPass)
rep$byTier
```

`rep` also carries `failures` (the failing checks, if any) and the full `checks`
list, and a `check_report.json` is written into `run_dir`. Use `rep$allPass` as a
gate in scripts.

## Optional: CDISC CORE conformance

Export SAS Transport (XPT v5) and, after running CORE against SDTMIG 3.4 + the
pinned CT, classify the findings:

```{r core}
export_xpt("runs/ref/sdtm", built$spec, "runs/ref/sdtm_xpt")  # haven::write_xpt, v5
# ...run CORE on the XPT, then:
digest_core("runs/ref/core_report/core.json",
            "runs/ref/core_report/summary.json")
```

`digest_core()` labels each finding as `data_bug` (fix the config and
regenerate), `tabulation_gap`, or `harness`.

## What domains can I generate?

A study populates only the domains its protocol implies; every other domain is
legitimately empty. Two helpers make scope explicit and auditable:

```{r domains}
supported_domains()   # domains generatable today (template + builder)
sdtm_domains()        # full 63-domain inventory with class + supported/findings_ready status
```

Findings-class domains reuse a single generic builder, so adding one usually means
adding a config test panel and a template entry — no new R builder. See the
package README for the full extension guide.

## Reproducibility in one line

```r
SDTM <- simulate_sdtm(config, spec, ct, n_subjects, seed)
```

Same inputs and seed ⇒ identical output. That contract is what lets `check_sdtm()`
recompute the expected shape from the config alone and assert on it.
